---
title: "Getting started with OMICstudio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with OMICstudio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What OMICstudio is

OMICstudio is a local, browser-based app for interactive multi-omics analysis.
You launch it with one command; it opens in your browser and runs all computation
on your own machine.

Five pipelines are offered on the start screen. The **single-cell RNA-seq**
(21 steps) and **WES / somatic mutation** (12 steps) pipelines are complete;
their cards are badged *Available*. Bulk RNA-seq, spatial transcriptomics and
multi-omics integration show their planned steps and are not implemented yet.

## Launch it

```{r}
OMICstudio::run_app()
```

This starts a local server and opens the app. Nothing is uploaded to any remote
server.

## Try it in 30 seconds (no data needed)

1. On the start screen, click the **Single-cell RNA-seq** card.
2. On the **1. Import** step, keep the source set to **Demo data**.
3. Leave the dataset as *PBMC 3k* — it is bundled in the package, so it loads
   instantly and offline — and click **Load demo data**.
4. Work down the left-hand navigator. Each step has a "What is this step?" card,
   a method choice, adjustable thresholds, and a large preview.

## Use your own data

On **1. Import**, choose **Upload file** and pick:

- an `.rds` Seurat / SingleCellExperiment object,
- a 10x `.h5`, or
- a counts table (CSV/TSV; genes in rows, cells in columns).

Old Seurat objects are updated automatically. You can also fetch a file from a
URL. Before doing anything else, the **Overview** tab surveys the raw data
(per-cell QC violins, top expressed genes, counts vs genes) so you can see what
you are about to filter.

## A typical run

Import → QC (adaptive MAD thresholds) → Doublet removal (scDblFinder) →
Normalize (LogNormalize) → Features + PCA → *(optional)* Integrate (Harmony) →
Cluster (Leiden) → Embed (UMAP) → Markers → Annotate (manual / SingleR /
Azimuth) → Enrichment → Trajectory → Visualize → Report → Export.

The **⤓ Export** menu in the top bar works on every step, so you can pull out the
object, metadata, counts or embeddings at any point. At **Export** you can also
download an R script that reproduces every step you ran, and at **Report** a
narrated HTML write-up of the same log.

## Linking your analysis to outcome

Step 18, **Clinical & survival**, takes a per-patient table (id, follow-up time,
outcome) and asks whether a grouping separates survival:

```{r}
# clinical.csv
# sample,os_months,os_status,stage
# P01,14.2,Dead,III
# P02,38.0,Alive,I
```

Stratify either by a column of that table, or by **cell composition** — the
fraction of a chosen cluster or cell type in each sample, split high vs low. You
get a Kaplan-Meier curve with the log-rank p-value, median survival per arm, and
a univariable Cox screen over any covariates you select.

Two things to know:

- Sample ids in the clinical table must match the object's sample metadata
  column for composition-based grouping; the app tells you how many matched.
- The **optimal** cutpoint is chosen to maximise separation, so its p-value is
  optimistic. Report the median split, or validate the cutpoint elsewhere.

The cohort is stored on the shared hub, so the WES pipeline picks it up too —
see below.

## The WES pipeline

Click the **WES / somatic mutations** card instead, and step 1 offers maftools'
bundled **TCGA LAML** cohort — 193 samples, instant and offline — so you can walk
the whole pipeline without data of your own.

```{r}
# what the app is doing under the hood
maf <- maftools::read.maf(
  system.file("extdata", "tcga_laml.maf.gz", package = "maftools"),
  clinicalData = read.delim(
    system.file("extdata", "tcga_laml_annot.tsv", package = "maftools")))
```

From there: the cohort summary, the oncoplot (annotate it by
`FAB_classification`), the mutation spectrum, TMB, a lollipop plot for any gene,
driver detection, signatures, clinical enrichment, a two-group comparison,
mutation-vs-survival and per-sample heterogeneity.

maftools installs as an ordinary Bioconductor binary, so unlike the single-cell
engine it needs no source build:

```{r}
BiocManager::install("maftools")
```

Only the signature step needs more: a `BSgenome` package for the reference build
and `NMF`. If your installed maftools has drifted from what the modules call,
`OMICstudio:::wes_missing_api()` lists exactly what is missing.

## Sharing one cohort between pipelines

`rv$clinical` is shared. Load a cohort in the single-cell **Clinical & survival**
step and the WES **Mutation vs survival** step offers it as a source — same
Kaplan-Meier code, same log-rank test, no second upload. The reverse works too:
the WES step writes the joined cohort back.

## Installing the heavy dependencies

The analysis engines (scop, Seurat, scDblFinder, harmony, SingleR, ...) are
optional at install time. If a step needs a package you don't have, the app tells
you what to install. To get everything at once without managing dependencies —
including scop and the conda environment for the Python-backed steps — use the
Docker image (see the README).
